Spanner: Google's Globally-Distributed Database*

### 2016-01-31 presentation/presentation_spanner.tex
Final version

\title[]{Spanner: Google's Globally-Distributed Database}
\subtitle[]{}

@@

\begin{document}
%-------------------------------------------------------------------------------
\begin{frame}
\titlepage
\end{frame}

@@

%-------------------------------------------------------------------------------
\section{Motivation}
\begin{frame}
	\frametitle{Motivation}
	\begin{itemize}
    \item{Distributed data at global scale}
    \item{Externally-consistent distributed transactions}
    \item{High availability}
    \item{Two "competitors": \emph{Bigtable} \& \emph{Megastore}}
    \pause
    \item{Why not use Bigtable?}
    \begin{itemize}
      \item{Difficult to use for applications with complex, evolving schemas}
      \item{Only \textbf{eventually-consistent} (no strong consistency)}
    \end{itemize}
    \pause
    \item{Why not use Megastore}
    \begin{itemize}
      \item{Poor write throughput}
    \end{itemize}

  \end{itemize}
\end{frame}

@@

%-------------------------------------------------------------------------------
\begin{frame}
  \frametitle{Fundamentals}
  \begin{itemize}
    \item{Transaction $i$ is denoted $T_i$}
    %\item{RW \ldots Read-Write (transaction)}
    %\item{RO \ldots Read-Only (transaction)}
    \item{\textbf{Externally-consistent:} \newline
      If $T_1$ preceeds $T_2$ (Commit of $T_1$ happens before Begin of $T_2$, no
      overlap), then $T_1$ is serialized first
    }
    \pause
    \item{\textbf{Two-Phase Locking (2PL):} \newline
    Guarantees serializability
    \begin{enumerate}
      \item{\emph{Expanding} phase: locks are aquired, no locks are released}
      \item{\emph{Shrinking} phase: locks are released, no locks are aquired}
    \end{enumerate}
    }
    \item{\textbf{Two-Phase Commit (2PC):} \newline
    Coordinates processes participating in atomic distributed transaction
    \begin{enumerate}
      \item{\emph{Commit-Request} phase: Request "Yes" (Commit) or "No" (Abort) from
        every transaction process
      }
      \item{\emph{Commit} phase: Commit transaction if all voted "Yes", otherwise abort}
    \end{enumerate}
    }
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}
  \frametitle{Features}
  As a globally-distributed database, Spanner provides interesting features:
  \begin{itemize}
    \item{Dynamically controlled replication configurations at fine grain}
    \item{Externally consistent reads \& writes}
    \item{globally-consistent reads at a timestamp}
  \end{itemize}
  $\Rightarrow$ enables \textbf{atomic schema changes} in presence of ongoing
  transactions \newline
  \pause
  \begin{itemize}
    \item{Timestamps reflect serialization order (linearizability)}
    \item{Key enabler: novel TrueTime API exposing clock uncertainty}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\section{Implementation}
\begin{frame}
  \frametitle{Implementation (Top-down)}
  \only<1-3>{
    \begin{itemize}
      \item{Spanner deployment is called \textbf{universe}}
      \item{Organized as a set of \textbf{zones}}
      \item{Zones are the unit of administrative deployment \& physical isolation}
      \item{Set of zones is the set of locations for replication}
      \pause
      \item{Each zone consists of}
      \begin{itemize}
        \item{several \textbf{spanservers} to store the data}
        \item{a \textbf{zonemaster} to assign data to spanservers}
        \item{a \textbf{location proxy} for clients to locate the assigned
          spanservers
        }
      \end{itemize}
      \pause
      \item{\textbf{universe master:} console to display status information about zones}
      \item{\textbf{placement driver:}}
      \begin{itemize}
        \item{handles automated data movement across zones}
        \item{moves data due to updated replication constraints or load balancing}
      \end{itemize}
    \end{itemize}
  }
  \only<4>{
    \begin{figure}
      \centering
      \includegraphics[width=.8\textwidth]{figures/spanner-server-organization.png}
      \caption{Spanner server organization}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Spanserver Software Stack}
\begin{frame}
  \frametitle{Spanserver Software Stack}
  \only<1>{
    \begin{itemize}
      \item{Responsible for 100 - 1000 \textbf{tablets}}
      \item{A tablet is a bag of key-value mappings \newline
        \texttt{(key:string, timestamp:int64) $\rightarrow$ string}
      }
      \item{Timestamps are assigned to data $\Rightarrow$
        \textbf{multi-version database}
      }
      \item{Tablet states are stored on Colossus}
      \item{Single Paxos state machine on top of each tablet $\Rightarrow$
        \textbf{Replication}
      }
      \item{Set of replicas is called \textbf{Paxos group}}
    \end{itemize}
  }
  \pause
  \only<2>{
    \begin{itemize}
      \item{Every replica which is a leader implements}
      \begin{itemize}
        \item{a \textbf{lock table} for concurrency control: \texttt{(key range) $\rightarrow$ lock state}}
        \item{a \textbf{transaction manager} (TM) for distributed transactions}
      \end{itemize}
      \item{Transaction manager used to implement \textbf{participant leader}}
      \item{If multiple Paxos groups are involved: Two-Phase Commit}
      \begin{itemize}
        \item{One participant group is chosen as \textbf{coordinator}}
        \item{Participant leader of this group: \textbf{coordinator leader}}
      \end{itemize}
      \item{State of each transaction manager is stored in Paxos group}
    \end{itemize}
  }
  \pause
  \only<3>{
    \begin{figure}
      \centering
      \includegraphics[width=.7\textwidth]{figures/spanserver-software-stack.png}
      \caption{Spanserver software stack}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Directories \& Placement}
\begin{frame}
  \frametitle{Directories \& Placement}
  \only<1-3>{
    \begin{itemize}
      \item{\textbf{Directories:}}
      \begin{itemize}
        \item{Bucketing abstraction on top of key-value mappings}
        \item{Contain set of contiguous keys sharing common prefix}
        \item{Control locality of data}
        \item{Unit of data placement (same replication configuration)}
        \item{Smallest unit to specify geographic-replication properties (\textbf{placement})}
      \end{itemize}
      \pause
      \item{Data is moved directory-wise between Paxos groups}
      \item{Tablets may have multiple \textbf{partitions} of row space \newline
        $\Rightarrow$ co-locate directories frequently accessed together
      }
      \pause
      \item{Placements can be controlled in two dimensions:}
      \begin{enumerate}
        \item{Number \& types of replicas}
        \item{Geographic placement of replicas}
      \end{enumerate}
      \item{\textbf{Movedir:} background task to move directories}
    \end{itemize}
  }
  \only<4>{
    \begin{figure}
      \centering
      \includegraphics[width=.6\textwidth]{figures/directories.png}
      \caption{Directories (the unit of data movement)}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Data Model}
\begin{frame}
  \frametitle{Data Model}
  \only<1>{
    \begin{itemize}
      \item{Features:}
      \begin{itemize}
        \item{Schematized semi-relational tables}
        \item{SQL-like query language}
        \item{General-purpose transactions}
      \end{itemize}
      \item{Layered on top of directory-bucketed key-value mappings}
      \item{Applications create \textbf{databases} in universe}
      \item{Database may contain unlimited number of schematized \textbf{tables}}
      \item{Table contains rows (must have names), columns \& versioned values}
    \end{itemize}
  }
  \only<2>{
    \begin{itemize}
      \item{Databases must be partitioned in one or more hierarchies}
      \item{\textbf{Directory table}: table at the top}
      \item{Directory contains ordered rows starting at a directory table key}
      \item{Interleaving tables to form directories is significant}
      \item{Locality relationships between multiple tables $\Rightarrow$ good
        performance
      }
    \end{itemize}

@@

%-------------------------------------------------------------------------------
\section{TrueTime}
\begin{frame}
  \frametitle{TrueTime}
  \only<1>{
    \begin{table}
      \centering
      \begin{tabular}{|l||l|}
        \hline
        Method & Returns \tabularnewline
        \hline\hline
        $TT.now()$ & $TTinterval$: [$earliest$, $latest$] \tabularnewline
        \hline
        $TT.after(t)$ & true if $t$ has definitely passed \tabularnewline
        \hline
        $TT.before(t)$ & true if $t$ has definitely not arrived \tabularnewline
        \hline
      \end{tabular}
      \caption{TrueTime API ($t$ is of type $TTstamp$)}
    \end{table}
    Time is represented as a $TTinterval$ (interval with bounded uncertainty)
    \newline\newline
    Let $e$ be an event and $t_{abs}(e)$ denote the absolute time of $e$.
    \newline
    For $tt = TT.now()$, the following property holds:
    \begin{center}
      $tt.earliest \leq t_{abs}(e) \leq tt.latest$
    \end{center}
  }
  \only<2>{
    \begin{itemize}
      \item{Time references: GPS and atomic clocks}
      \item{Set of \textbf{time master} machines per datacenter}
      \item{\textbf{Armageddon masters}: masters using atomic clocks}
      \item{\textbf{Timeslave daemon} on each machine}
      \pause
      \item{Masters compare their time references regularly}
      \item{Daemons poll a variety of masters (liar detection)}
      \pause
      \item{$\epsilon$ is the instantaneous error bound (typically 1 - 7ms)}
      \item{$\overline{\epsilon}$ is the average error bound (typically 4ms)}
      \item{$\epsilon$ is derived from applied worst-case local clock drift ($200\mu s/sec$)}
    \end{itemize}
  }
  \only<3>{
    \begin{figure}
      \centering
      \begin{tikzpicture}[thick]
        \node (left-end) at (0, 0) {};
        \node[right = 6 of left-end] (right-end) {};

@@

%-------------------------------------------------------------------------------
\section{Concurrency Control}
\begin{frame}
  \frametitle{Concurrency Control}
  \begin{itemize}
    \item{TrueTime is used to guarantee correctness properties}
    \item{Properties are used to implement}
    \begin{itemize}
      \item{Externally-consistent transactions}
      \item{Lock-free read-only transactions}
      \item{Non-blocking reads in the past}
    \end{itemize}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Timestamp Management}
\begin{frame}
  \frametitle{Timestamp Management}
  \only<1>{
    \footnotesize
    \begin{table}
      \begin{tabular}{|l||l|l|}
        \hline
        Operation & CC & Replica required\tabularnewline
        \hline\hline
        Read-Write transaction & Pessimistic & leader\tabularnewline
        \hline
        Read-Only transaction & Lock-free & leader; any
        \tabularnewline
        \hline
        Snapshot Read, client-provided timestamp & Lock-free & any\tabularnewline
        \hline
        Snapshot Read, client-provided bound & Lock-free & any \tabularnewline
        \hline
      \end{tabular}
      \caption{Supported types of reads \& writes}
    \end{table}
  }
  \begin{itemize}
    \item{Standalone write $\Rightarrow$ RW transaction}
    \item{Non-snapshot standalone read $\Rightarrow$ RO transaction}
    \item{RO transactions}
    \begin{itemize}
      \item{must predeclare to not include writes}
      \item{can proceed on any sufficiently up-to-date replica (also snapshot
        reads)
      }
    \end{itemize}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Paxos Leader Leases}
\begin{frame}
  \frametitle{Paxos Leader Leases}
  \begin{itemize}
    \item{Timed leases for long-lived leadership (10s by default)}
    \item{Quorum of lease votes $\Rightarrow$ leader has a lease}
    \item{Successful writes extend lease vote on replica}
    \item{\textbf{Disjointness invariant:} \newline
      For each Paxos group, each Paxos leader's \emph{lease interval} is disjoint
      from every other leader's
    }
    \item{Leaders must not abdicate before $TT.after(s_{max})$, where $s_{max}$
      denotes the maximum timestamp used by a leader
    }
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Timestamps in RW Transactions}
\begin{frame}
  \frametitle{Timestamps in RW Transactions}
  \only<1>{
    \begin{itemize}
      \item{Timestamp assignment only in between the two phases (2PL)}
      \item{\textbf{Monotonicity variant:} \newline
        Within each Paxos group, Spanner assigns timestamps to Paxos writes in
        monotonically increasing order, even across leaders
      }
      \item{Enforced across leaders by disjointness invariant: \newline
        Leader must only assign timestamps within the interval of leader lease
      }
      \item{Timestamp $s$ assigned $\Rightarrow$ $s_{max} = s$}
      \item{\textbf{External consistency invariant}: \newline
        \begin{center}
          $t_{abs}\left(e^{commit}_1\right) < t_{abs}\left(e^{start}_2\right)$
          $\Rightarrow s_1 < s_2$
        \end{center}
      }
    \end{itemize}
  }
  \only<2>{
    \begin{itemize}
      \item{Let $e^{server}_i$ denote the arrival event of commit request for
        $T_i$
      }
      \item{Two rules guarantee the external consistency invariant:}
      \begin{enumerate}
        \item{\textbf{Start} \newline
          Coordinator leader assigns commit timestamp \newline
          $s_i \geq TT.now().latest$ after $e^{server}_i$
        }
        \item{\textbf{Commit Wait} \newline
          Coordinator leader ensures clients cannot see \newline
          data commited by $T_i$ until $TT.after(s_i)$ is true
        }
      \end{enumerate}
    \end{itemize}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Serving Reads at a Timestamp}
\begin{frame}
  \frametitle{Serving Reads at a Timestamp}
  \begin{itemize}
    \item{Monotonicity invariant allows to determine if a replica's state is
      \textbf{sufficiently up-to-date} to satisfy a read
    }
    \item{\textbf{Safe time} $t_{safe}$ at each replica}
    \item{Replica satisfies read if $t \leq t_{safe}$}
    \item{$t_{safe} = min\left(t^{Paxos}_{safe}, t^{TM}_{safe}\right)$}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Timestamps in RO Transactions}
\begin{frame}
  \frametitle{Timestamps in RO Transactions}
  \begin{itemize}
    \item{Two phases:}
    \begin{enumerate}
      \item{Assign timestamp $s_{read}$}
      \item{Execute transaction's reads as snapshot reads at $s_{read}$}
    \end{enumerate}
    \item{Spanner assigns the oldest timestamp that preserves external consistency
      to reduce the changes of blocking
    }
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Details}

\subsubsection{Read-Write Transactions}
\begin{frame}
  \frametitle{Read-Write Transactions}
  \only<1-3>{
    \begin{itemize}
      \item{RW transactions buffered at client until Commit}
      \item{Reads do not see effects of transaction's writes}
      \item{Client has completed all reads \& buffered all writes $\Rightarrow$ 2PC}
      \pause
      \item{Non-coordinator-participant leader}
      \begin{enumerate}
        %\item{Acquires write locks}
        \item{Chooses prepare timestamp according to monotonicity}
        \item{Logs prepare record through Paxos}
        \item{Notifies coordinator of its prepare timestamp}
      \end{enumerate}
      \pause
      \item{Coordinator leader}
      \begin{enumerate}
        %\item{Aquires write locks}
        \item{Chooses timestamp for whole transaction}
        \item{Commit timestamp $s$ must be}
        \begin{itemize}
          \item{$\geq$ all prepare timestamps}
          \item{$> TT.now().latest$ at the time the Commits were received}
          \item{$>$ any previously assigned timestamp}
        \end{itemize}
        \item{Logs Commit record through Paxos}
      \end{enumerate}
    \end{itemize}
  }
  \only<4-9>{
    \begin{figure}
      \centering
      \begin{tikzpicture}[thick]
        \node (left-end-t1) at (0,2) {};
        \node[right = 9 of left-end-t1] (right-end-t1) {};
        \draw[ultra thick, ->, >=stealth] (left-end-t1.center) -- (right-end-t1.center)
          node[right] (right-end-t1-t) {$t$};
          \node[left = 0 of left-end-t1.west] (left-end-t1-t) {$\mathbf{T_C}$};

        \node[below = 1 of left-end-t1] (left-end-t2) {};
        \node[right = 9 of left-end-t2] (right-end-t2) {};
        \draw[->, >=stealth] (left-end-t2.center) -- (right-end-t2.center)
          node[right] (right-end-t2-t) {$t$};
        \node[left = 0 of left-end-t2.west] (left-end-t2-t) {$T_2$};

        \node[below = 1 of left-end-t2] (left-end-t3) {};
        \node[right = 9 of left-end-t3] (right-end-t3) {};
        \draw[->, >=stealth] (left-end-t3.center) -- (right-end-t3.center)
          node[right] (right-end-t3-t) {$t$};
        \node[left = 0 of left-end-t3.west] (left-end-t3-t) {$T_3$};

>>>>

\title[]{Spanner: Google's Globally-Distributed Database*}
\subtitle[]{}

@@

\begin{document}
%-------------------------------------------------------------------------------
\begin{frame}
\titlepage
{\footnotesize* James C. Corbett et al. (26 authors), OSDI 2012.}
\end{frame}

@@

%-------------------------------------------------------------------------------
\section{Motivation}
\begin{frame}
	\frametitle{Motivation}
	\begin{itemize}
    \item{Distributed data at global scale}
    \item{Externally-consistent distributed transactions}
    \item{High availability}
    \pause
    \item{Why not use Bigtable?}
    \begin{itemize}
      \item{Difficult to use for applications with complex, evolving schemas}
      \item{Only \textbf{eventually-consistent} (no strong consistency)}
    \end{itemize}
    \pause
    \item{Why not use Megastore?}
    \begin{itemize}
      \item{Poor performance}
    \end{itemize}
  \end{itemize}
\end{frame}

@@

%-------------------------------------------------------------------------------
\begin{frame}
  \frametitle{Fundamentals}
  \only<1>{
    \begin{itemize}
      %\item{Transaction $i$ is denoted $T_i$}
      %\item{RW \ldots Read-Write (transaction)}
      %\item{RO \ldots Read-Only (transaction)}
      \item{\textbf{External consistency:} \newline
        If $T_1$ preceeds $T_2$, then $T_1$ is serialized first
      }
      \item{\textbf{Paxos (Replication):}
        \begin{itemize}
          \item{Solves problem of resilient replication of data}
          \item{Data eventually propagates to all nodes}
          \item{Different nodes always see the same data}
          \item{Majority of nodes up $\Rightarrow$ Writes/Reads processed
            correctly
          }
          \item{Single node is elected as leader and initiates consensus}
          \item{Guarantees \textbf{consistency}}
        \end{itemize}
      }
    \end{itemize}
  }
  \only<2>{
    \begin{itemize}
      \item{\textbf{Two-Phase Locking (2PL):} \newline
      Guarantees serializability
      \begin{enumerate}
        \item{\emph{Expanding} phase: locks are aquired, no locks are released}
        \item{\emph{Shrinking} phase: locks are released, no locks are aquired}
      \end{enumerate}
      }
      \item{\textbf{Two-Phase Commit (2PC):} \newline
      Coordinates processes participating in atomic distributed transaction
      \begin{enumerate}
        \item{\emph{Commit-Request} phase: Request "Yes" (Commit) or "No" (Abort) from
          every transaction process
        }
        \item{\emph{Commit} phase: Commit transaction if all voted "Yes", otherwise abort}
      \end{enumerate}
      }
    \end{itemize}
  }
\end{frame}

%-------------------------------------------------------------------------------
\begin{frame}
  \frametitle{Features}
  As a globally-distributed database, Spanner provides interesting features:
  \begin{itemize}
    \item{Dynamically controlled replication configurations at fine grain}
    \item{Externally consistent reads \& writes}
    \item{globally-consistent reads at a timestamp}
  \end{itemize}
  $\Rightarrow$ enables \textbf{atomic schema changes} in presence of ongoing
  transactions \newline
  \begin{itemize}
    \item{Timestamps reflect serialization order}
    \item{Key enabler: novel \textbf{TrueTime} API \textbf{exposing clock uncertainty}}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\section{Implementation}
\begin{frame}
  \frametitle{Implementation}
  %\only<1-3>{
  %  \begin{itemize}
  %    \item{Spanner deployment is called \textbf{universe}}
  %    \item{Organized as a set of \textbf{zones}}
  %    \item{Zones are the unit of administrative deployment \& physical isolation}
  %    \item{Set of zones is the set of locations for replication}
  %    \pause
  %    \item{Each zone consists of}
  %    \begin{itemize}
  %      \item{several \textbf{spanservers} to store the data}
  %      \item{\textbf{location proxies} for clients to locate the assigned
  %        spanservers
  %      }
  %      \item{a \textbf{zonemaster} to assign data to spanservers}
  %    \end{itemize}
  %    \pause
  %    \item{\textbf{universe master:} console to display status information about zones}
  %    \item{\textbf{placement driver:}}
  %    \begin{itemize}
  %      \item{handles automated data movement across zones}
  %      \item{moves data due to updated replication constraints or load balancing}
  %    \end{itemize}
  %  \end{itemize}
  %}
  %\only<4>{
    \begin{figure}
      \centering
      \includegraphics[width=.8\textwidth]{figures/spanner-server-organization.png}
      \caption{Spanner server organization}
    \end{figure}
  %}
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Spanserver Software Stack}
\begin{frame}
  \frametitle{Spanserver Software Stack}
  \only<1>{
    \begin{itemize}
      \item{Responsible for 100 - 1000 \textbf{tablets}}
      \item{A tablet is a bag of key-value mappings \newline
        \texttt{(key:string, timestamp:int64) $\rightarrow$ string}
      }
      \item{Timestamps are assigned to data $\Rightarrow$
        \textbf{multi-version database}
      }
      \item{Tablet states are stored on Colossus}
      \item{Single Paxos state machine on top of each tablet $\Rightarrow$
        \textbf{Replication}
      }
      \item{Set of replicas is called \textbf{Paxos group}}
    \end{itemize}
  }
  \pause
  \only<2>{
    \begin{itemize}
      \item{Every replica which is a leader implements}
      \begin{itemize}
        \item{a \textbf{lock table} for concurrency control: state for 2PL}
        \item{a \textbf{transaction manager} (TM) for distributed transactions}
      \end{itemize}
      \item{Transaction manager used to implement \textbf{participant leader}}
      \item{If multiple Paxos groups are involved: Two-Phase Commit}
      \begin{itemize}
        \item{One participant group is chosen as \textbf{coordinator group}}
        \item{Participant leader of this group: \textbf{coordinator leader}}
      \end{itemize}
      \item{State of each transaction manager is stored in Paxos group}
    \end{itemize}
  }
  \pause
  \only<3>{
    \begin{figure}
      \centering
      \includegraphics[width=.7\textwidth]{figures/spanserver-software-stack.png}
      \caption{Spanserver software stack}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Directories \& Placement}
\begin{frame}
  \frametitle{Directories \& Placement}
  \only<1-2>{
    \begin{itemize}
      \item{\textbf{Directories:}}
      \begin{itemize}
        \item{Bucketing abstraction on top of tablet}
        \item{Contain set of contiguous keys sharing common prefix}
        \item{Control locality of data (pyhsical)}
        \item{Unit of data placement (same replication configuration)}
        \item{Smallest unit to specify geographic-replication properties
          (\textbf{placement})
        }
      \end{itemize}
      \pause
      \item{Data is moved directory-wise between Paxos groups}
      \item{Tablets may have multiple directories \newline
        $\Rightarrow$ co-locate data frequently accessed together
      }
      \item{Placements can be controlled in two dimensions:}
      \begin{enumerate}
        \item{Number \& types of replicas}
        \item{Geographic placement of replicas}
      \end{enumerate}
      % \item{\textbf{Movedir:} background task to move directories}
    \end{itemize}
  }
  \only<3>{
    \begin{figure}
      \centering
      \includegraphics[width=.6\textwidth]{figures/directories.png}
      \caption{Directories (the unit of data movement)}
    \end{figure}
  }
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Data Model}
\begin{frame}
  \frametitle{Data Model}
  \only<1>{
    \begin{itemize}
      %\item{Features:}
      %\begin{itemize}
      %  \item{Schematized semi-relational tables}
      %  \item{SQL-like query language}
      %  \item{General-purpose transactions}
      %\end{itemize}
      \item{Layered on top of directory-bucketed key-value mappings}
      \item{Applications create \textbf{databases}}
      \item{Database may contain any number of schematized \textbf{tables}}
      \item{Table contains rows, columns \& versioned values}
    \end{itemize}
  }
  \only<2>{
    \begin{itemize}
      \item{Databases must be partitioned in one or more hierarchies}
      \item{\textbf{Directory table}: table at the top}
      \item{Directory contains hierarchically ordered rows}
      \item{Interleaving tables $\Rightarrow$ locality relationships of multiple
        tables \newline$\Rightarrow$ significant for performance
      }
    \end{itemize}

@@

%-------------------------------------------------------------------------------
\section{TrueTime}
\begin{frame}
  \frametitle{TrueTime}
  \only<1>{
    \begin{center}
      {\footnotesize
      \begin{tabular}{|l||l|}
        \hline
        Method & Returns \tabularnewline
        \hline\hline
        $TT.now()$ & $TTinterval$: [$earliest$, $latest$] \tabularnewline
        \hline
        $TT.after(t)$ & true if $t$ has definitely passed \tabularnewline
        \hline
        $TT.before(t)$ & true if $t$ has definitely not arrived \tabularnewline
        \hline
      \end{tabular}
      }
    %  \caption{TrueTime API ($t$ is of type $TTstamp$)}
    \end{center}
    Time is represented as a $TTinterval$ (interval with bounded uncertainty)
    \newline\newline
    %Let $e$ be an event and $t_{abs}(e)$ denote the absolute time of $e$.
    %\newline
    %For $tt = TT.now()$, the following property holds:
    %\begin{center}
    %  $tt.earliest \leq t_{abs}(e) \leq tt.latest$
    %\end{center}
    \textbf{Guarantee:} $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$
  }
  \only<2>{
    \begin{itemize}
      \item{Time references: GPS and atomic clocks}
      \item{Set of \textbf{time master} machines per datacenter}
      %\item{\textbf{Armageddon masters}: masters using atomic clocks}
      \item{\textbf{Timeslave daemon} on each machine}
      \item{Masters compare time references regularly}
      \item{Daemons poll a variety of masters}
      \item{$\epsilon$ is the \textbf{instantaneous error bound} (typically 1 -
        7ms)
      }
      \item{$\overline{\epsilon}$ is the average error bound (typically 4ms)}
      \item{$\epsilon$ is derived from applied worst-case local clock drift ($200\mu s/sec$)}
    \end{itemize}
  }
  \only<3>{
    \begin{figure}
      \centering
      \begin{tikzpicture}[thick]
        \node (left-end) at (0, 0) {};
        \node[right = 6 of left-end] (right-end) {};

@@

%-------------------------------------------------------------------------------
\section{Concurrency Control}
\begin{frame}
  \frametitle{Concurrency Control}
  \begin{itemize}
    \item{TrueTime used to guarantee correctness properties to implement:}
    \begin{itemize}
      \item{Externally-consistent transactions}
      \item{Lock-free read-only transactions}
      \item{Non-blocking reads in the past (snapshot reads)}
    \end{itemize}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsection{Timestamp Management}
\begin{frame}
  \frametitle{Timestamp Management}
  \only<1>{
    \begin{center}
      {\footnotesize
      \begin{tabular}{|l||l|l|}
        \hline
        Operation & CC & Replica required\tabularnewline
        \hline\hline
        Read-Write transaction & Pessimistic & leader\tabularnewline
        \hline
        Read-Only transaction & Lock-free & leader; any
        \tabularnewline
        \hline
        Snapshot Read, client-provided timestamp & Lock-free & any\tabularnewline
        \hline
        Snapshot Read, client-provided bound & Lock-free & any \tabularnewline
        \hline
      \end{tabular}
      }
    %  \caption{Supported types of reads \& writes}
    \end{center}
  }
  \begin{itemize}
    \item{Standalone write $\Rightarrow$ Read-Write (RW) transaction}
    \item{Non-snapshot standalone read $\Rightarrow$ Read-Only (RO) transaction}
    \item{RO transactions must predeclare to not include writes}
    \item{RO transactions \& snapshot reads proceed on any
      \textbf{sufficiently up-to-date} replica
    }
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Paxos Leader Leases}
\begin{frame}
  \frametitle{Paxos Leader Leases}
  \begin{itemize}
    \item{Timed leases for long-lived leadership (10s by default)}
    \item{Quorum of lease votes $\Rightarrow$ leader has a lease}
    \item{Successful writes extend lease vote on replica}
    \item{\textbf{Disjointness invariant:} \newline
      For each Paxos group, each Paxos leader's \emph{lease interval} is disjoint
      from every other leader's
    }
    \item{Leaders must not abdicate before $TT.after(s_{max})$}
    \item{$s_{max}$ \ldots maximum timestamp used by a leader}
  \end{itemize}
\end{frame}

%-------------------------------------------------------------------------------
\subsubsection{Read-Write Transactions}
\begin{frame}
  \frametitle{Read-Write Transactions}
  \only<1>{
    \begin{itemize}
      \item{Uses 2PL: Timestamp assignment in between the two phases}
      \item{\textbf{Monotonicity invariant:} \newline
        Within each Paxos group, Spanner assigns timestamps to Paxos writes in
        monotonically increasing order, even across leaders
      }
      \item{Enforced across leaders by disjointness invariant: \newline
        Leader must only assign timestamps within interval of leader lease
      }
      \item{Timestamp $s$ assigned $\Rightarrow$ $s_{max} = s$}
      \item{\textbf{External consistency invariant}: \newline
        \begin{center}
          $t_{abs}\left(e^{commit}_1\right) < t_{abs}\left(e^{start}_2\right)$
          $\Rightarrow s_1 < s_2$
        \end{center}
      }
    \end{itemize}
  }
  \only<2>{
    \begin{itemize}
      \item{Writes buffered at client until commit} 
      \item{Client has completed all reads \& buffered all writes $\Rightarrow$ 2PC}
      \item{Let $e^{server}_i$ denote the arrival event of commit request for
        $T_i$
      }
      \item{Two rules guarantee the external consistency invariant:}
      \begin{enumerate}
        \item{\textbf{Start} \newline
          Coordinator leader assigns commit timestamp \newline
          $s_i \geq TT.now().latest$ after $e^{server}_i$
        }
        \item{\textbf{Commit Wait} \newline
          Coordinator leader ensures clients cannot see \newline
          data commited by $T_i$ until $TT.after(s_i)$ is true

        }
      \end{enumerate}
      %\pause
      %\item{Non-coordinator-participant leader}
      %\begin{enumerate}
      %  \item{Acquires write locks}
      %  \item{Chooses prepare timestamp according to monotonicity}
      %  \item{Logs prepare record through Paxos}
      %  \item{Notifies coordinator of its prepare timestamp}
      %\end{enumerate}
      %\pause
      %\item{Coordinator leader}
      %\begin{enumerate}
      %  \item{Aquires write locks}
      %  \item{Chooses timestamp for whole transaction}
      %  \item{Commit timestamp $s$ must be}
      %  \begin{itemize}
      %    \item{$\geq$ all prepare timestamps}
      %    \item{$> TT.now().latest$ at the time the commits were received}
      %    \item{$>$ any previously assigned timestamp}
      %  \end{itemize}
      %  \item{Logs Commit record through Paxos}
      %\end{enumerate}
    \end{itemize}
  }
  \only<3-8>{
    \begin{figure}
      \centering
      \begin{tikzpicture}[thick]
        \node (left-end-t1) at (0,2) {};
        \node[right = 9 of left-end-t1] (right-end-t1) {};
        \draw[ultra thick, ->, >=stealth] (left-end-t1.center) -- (right-end-t1.center)
          node[right] (right-end-t1-t) {$t$};
          \node[left = 0 of left-end-t1.west] (left-end-t1-t) {$\mathbf{P_{CL}}$};

        \node[below = 1 of left-end-t1] (left-end-t2) {};
        \node[right = 9 of left-end-t2] (right-end-t2) {};
        \draw[->, >=stealth] (left-end-t2.center) -- (right-end-t2.center)
          node[right] (right-end-t2-t) {$t$};
        \node[left = 0 of left-end-t2.west] (left-end-t2-t) {$P_2$};

        \node[below = 1 of left-end-t2] (left-end-t3) {};
        \node[right = 9 of left-end-t3] (right-end-t3) {};
        \draw[->, >=stealth] (left-end-t3.center) -- (right-end-t3.center)
          node[right] (right-end-t3-t) {$t$};
        \node[left = 0 of left-end-t3.west] (left-end-t3-t) {$P_3$};
